A Mechanised Library for Generic (Co)Programming

### 2024-03-06 paper/paper.tex
working on recursive coalgebras

\subsubsection{On Equality}
This definition highlights why we cannot define equality of the positions and
the contained elements generically (recall Section~\ref{sec:TODO}): the
contained types may not have (yet) an associated equivalence. In other words, we
would need to provide an equivalence for \coq{LFix}, but the only way to do this
is to define such equivalence simultaneously to the definition for  \coq{LFix}.
Such definition would be rejected by Coq. There are other approaches that can
circumvent this problem, by separating the definition of least fixed points from
the proofs that the functions that get the contained elements are proper
morphisms. But this leads to longer, unnecessarily complicated proofs.
Furthermore, this extra layer interferes with code extraction, in that it
requires destructing/constructing subsequent \coq{LFix} and \coq{App}, and Coq
may not always able to inline them.

\subsection{Recursive Coalgebras and Hylomorphism}
\label{sec:rec-coalgebras}

Hylomorphism are more general recursion schemes capturing the
\emph{divide-and-conquer} idea. In short, the input is destructured by its not
structurally. In the example, we gave in the introduction the destructuring
program is given by \coq{divide} which is not an isomorphism.  The question the
hylomorphism satisfies is more general than the catamorphism in that the
destructing map \coq{InOp} is generalised to a \emph{recursive
coalgebra}~\cite{AdamekMM19,CaprettaUV04}, that is intuitively a coalgebra that
can be applied only a finite amount of times.  In fact, in the case
of \coq{divide} this is a recursive coalgebra because the size of the list is
getting smaller.

% This program is formed of essentially two parts. The first one is what is
% commonly thought as the \emph{partition} function destructuring the input list
% in three parts, the pivot and the other two sublists, and the second part
% recursing over the sublists and composing the results back together.

In particular, we show that the terminal coalgebras
for $F$, coupled with a proof that they are finite are also initial
$F$-algebras:
\begin{coqcode}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{coqcode}
Catamorphisms from this initial algebra are denoted by \coq{ccata},
and they can be used to define the isomorphism between
\coq{{x : GFix F | FinF x}} and
\coq{LFix F}. In particular:
\begin{coqcode}
Lemma cata_ccata `{setoid A} (f : Alg F A) : cata f \o ccata l_in =e ccata f.
Lemma ccata_cata `{setoid A} (f : Alg F A) : ccata f \o cata lg_in =e cata f.
\end{coqcode}
Using this new initial algebra, we can compose catamorphisms and anamorphisms,
\emph{as long as the anamorphisms produce finite trees}. We refer the reader to
the supplementary material and Coq code to see how we encode this. In this
paper, however, we focus on the fact that, if anamorphisms produce finite
trees, then their return type can also be the \emph{least} fixed point of the
container extension. We encode in Coq the notion of \emph{recursive
anamorphisms} and \emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}.  Recursive
coalgebras are coalgebras whose anamorphism produces finite trees.  The
following predicate captures this fact, \coq{h} is recursive if
\coq{h} can only be applied finitely many times:
\begin{coqcode}
Inductive RecF (h : Coalg F A) : A -> Prop :=
| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
\end{coqcode}
This predicate is equivalent to the proposition that the anamorphism of such
coalgebras produce only finite trees:
\begin{coqcode}
Lemma ana_rec_term `{setoid A} (h : Coalg F A)
  : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
\end{coqcode}
Recursive anamorphisms are implemented in Coq recursively on the structure of
the proof that they can only be applied finitely many times (\coq{RecF}).  Note
that such proofs will be erased when extracting code to OCaml, and this is
standard practice when building general recursive functions in
Coq\footnote{\dcas{This technique had a name in the community, but I cannot
remember. Something predicates}}. The core definitions are the following:
\begin{coqcode}
Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).

@@

\subsubsection{Hylomorphisms}

Given an $F$-algebra and $F$-coalgebra, a solution to the equation
$f = a \circ F\; f \circ c$ is known as a hylomorphism. Unfortunately, this
definition cannot be accepted by Coq, since there is no structurally smaller
argument to do recursion, and there is no corecursive definition that is
guaranteed to be syntactically guarded (as required by Coq), due to the use of
an arbitrary algebra $a$. Furthermore, solutions to the hylomorphism equation
are \emph{not} guaranteed to be unique, or to even exist. To address these
issues, we mechanise \textbf{recursive hylomorphisms} which are guaranteed to
have a unique solution to the hylomorphism equation.

If $c$ is a recursive coalgebra, then the equation $f = a \circ F\; f \circ c$
does have a unique solution.  We show this by first providing the definition for
recursive hylomorphisms in Coq, and then proving that any function satisfying
this equation is a hylomorphism. We call these hylomorphisms with a recursive
coalgebra, \textbf{recursive hylomorphisms}.
\begin{coqcode}
Definition hylo_def (a : Alg F B) (c : Coalg F A)
  : forall (x : A), RecF c x -> B := fix f x H
  := match c x as h0 return
       (forall e : Pos (shape h0), RecF c (cont h0 e)) -> B
     with
     | MkApp s_x c_x => fun H => a (MkApp s_x (fun e => f (c_x e) (H e)))
     end (RecF_inv H).
\end{coqcode}
The definition of recursive hylomorphisms is recursive on the proof that the
anamorphism can only be applied finitely many times on any input. We again use
\coq{RecF_inv} to obtain the structurally smaller proof to use in
the recursive calls.  As we did with catamorphisms and recursive anamorphisms,
we prove that \coq{hylo_def} is respectful, and use this proof to
built the corresponding higher-order proper morphism:
\begin{coqcode}
Definition hylo : Alg F B ~> RCoalg F A ~> A ~> B := ...
\end{coqcode}
Finally, we show that recursive hylomorphisms are the unique solution to
the hylomorphism equation.
\begin{coqcode}
Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
    : f =e hylo g h <-> f =e g \o fmap f \o h.
\end{coqcode}
This result can be used to prove hylomorphism laws in a way that is analogous
to pen-and-paper program calculation techniques. We show the rule for
hylomorphism fusion as an example:
\begin{coqcode}
Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C)
  (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2) : f2 \o hylo g1 h1 =e hylo g2 h1.

@@

>>>>

\subsection{Recursive Coalgebras and Hylomorphism}
\label{sec:rec-coalgebras}
Hylomorphism capture the concept of \emph{divide-and-conquer} algorithms.  First
the input is destructured (\emph{divide}) in smaller parts. This operation may or may not be
structural. Next the results from the subparts are computed recursively and
finally the results are composed together (\emph{conquer}) by means of an algebra.

A catamorphism is a special kind of hylomorphism where the coalgebra
map \coq{InOp} only destructures the input structurally, for example,
decomposing a list into its head and tail. A \emph{recursive
coalgebra}~\cite{AdamekMM19,CaprettaUV04} is a coalgebra which does not
necessarily decompose the input structurally, still maintaining the size of the
input smaller. This is what, for example, happens in quicksort where the
partition function, given a list, returns two smaller sublists and a pivot
element.

Thus, in general given an $F$-algebra and $F$-coalgebra, the hylomorphism is the unique solution to the equation
\begin{equation}
  \label{eq:hylo}
  f = a \circ F\; f \circ c
\end{equation}
However this solution does not necessarily exists for an arbitrary pair of
algebras and coalgebras.  In fact, this definition cannot be accepted by Coq,
since there is no structurally smaller argument to do recursion on and there is
no corecursive definition that is guaranteed to be syntactically guarded -- as
required by Coq -- due to the use of an arbitrary algebra $a$. Furthermore,
solutions to the hylomorphism equation are \emph{not} guaranteed to be unique,
or to even exist. To address these issues, we mechanise \textbf{recursive
hylomorphisms} which are guaranteed to have a unique solution to the
hylomorphism equation.

If $c$ is a recursive coalgebra, then the equation (\ref{eq:hylo}) has a unique
solution. We show this by first providing the definition for recursive
hylomorphisms in Coq, and then proving that any function satisfying this
equation is a hylomorphism.
\begin{coqcode}
Definition hylo_def (a : Alg F B) (c : Coalg F A)
  : forall (x : A), RecF c x -> B := fix f x H
  := match c x as h0 return
       (forall e : Pos (shape h0), RecF c (cont h0 e)) -> B
     with
     | MkApp s_x c_x => fun H => a (MkApp s_x (fun e => f (c_x e) (H e)))
     end (RecF_inv H).
\end{coqcode}
The definition of recursive hylomorphisms is recursive on the proof that the
anamorphism can only be applied finitely many times on any input. We again use
\coq{RecF_inv} to obtain the structurally smaller proof to use in
the recursive calls.  As we did with catamorphisms and recursive anamorphisms,
we prove that \coq{hylo_def} is respectful, and use this proof to
built the corresponding higher-order proper morphism:
\begin{coqcode}
Definition hylo : Alg F B ~> RCoalg F A ~> A ~> B := ...
\end{coqcode}
Finally, we show that recursive hylomorphisms are the unique solution to
the hylomorphism equation.
\begin{coqcode}
Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
    : f =e hylo g h <-> f =e g \o fmap f \o h.
\end{coqcode}
This result can be used to prove hylomorphism laws in a way that is analogous
to pen-and-paper program calculation techniques. We show the rule for
hylomorphism fusion as an example:
\begin{coqcode}
Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C)
  (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2) : f2 \o hylo g1 h1 =e hylo g2 h1.

\mpav{Moved here}
In particular, we show that the terminal coalgebras
for $F$, coupled with a proof that they are finite are also initial
$F$-algebras:
\begin{coqcode}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{coqcode}
Catamorphisms from this initial algebra are denoted by \coq{ccata},
and they can be used to define the isomorphism between
\coq{{x : GFix F | FinF x}} and
\coq{LFix F}. In particular:
\begin{coqcode}
Lemma cata_ccata `{setoid A} (f : Alg F A) : cata f \o ccata l_in =e ccata f.
Lemma ccata_cata `{setoid A} (f : Alg F A) : ccata f \o cata lg_in =e cata f.
\end{coqcode}
Using this new initial algebra, we can compose catamorphisms and anamorphisms,
\emph{as long as the anamorphisms produce finite trees}. We refer the reader to
the supplementary material and Coq code to see how we encode this. In this
paper, however, we focus on the fact that, if anamorphisms produce finite trees,
then their return type can also be the \emph{least} fixed point of the container
extension. We encode in Coq the notion of \emph{recursive anamorphisms} and
\emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}. Recursive
coalgebras are coalgebras whose anamorphism produces finite trees. The
following predicate captures this fact, \coq{h} is recursive if \coq{h} can only
be applied finitely many times:
\begin{coqcode}
Inductive RecF (h : Coalg F A) : A -> Prop :=
| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
\end{coqcode}
This predicate is equivalent to the proposition that the anamorphism of such
coalgebras produce only finite trees:
\begin{coqcode}
Lemma ana_rec_term `{setoid A} (h : Coalg F A)
  : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
\end{coqcode}
Recursive anamorphisms are implemented in Coq recursively on the structure of
the proof that they can only be applied finitely many times (\coq{RecF}).  Note
that such proofs will be erased when extracting code to OCaml, and this is
standard practice when building general recursive functions in
Coq\footnote{\dcas{This technique had a name in the community, but I cannot
remember. Something predicates}}. The core definitions are the following:
\begin{coqcode}
Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).

@@

@@

\section{On Equality}
This definition highlights why we cannot define equality of the positions and
the contained elements generically (recall Section~\ref{sec:TODO}): the
contained types may not have (yet) an associated equivalence. In other words, we
would need to provide an equivalence for \coq{LFix}, but the only way to do this
is to define such equivalence simultaneously to the definition for  \coq{LFix}.
Such definition would be rejected by Coq. There are other approaches that can
circumvent this problem, by separating the definition of least fixed points from
the proofs that the functions that get the contained elements are proper
morphisms. But this leads to longer, unnecessarily complicated proofs.
Furthermore, this extra layer interferes with code extraction, in that it
requires destructing/constructing subsequent \coq{LFix} and \coq{App}, and Coq
may not always able to inline them.
